An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems

### 2014-04-30 paper/paper.tex
Paper: finish incorporating suggestions from mihaly in Background

% TODO make title page look nicer
\begin{titlepage}
\begin{center}
\thispagestyle{empty}
% TODO install \textsc font
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[1.75cm]
\LARGE{\textsc {\textbf{Improving the Speed of Peer to Peer Backup Systems with BitTorrent}}}\\[0.5cm]
\vspace{0.25cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.75cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{1cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\vspace{.25cm}
\large{\textbf{Mih\'aly H\'eder}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\end{center}
\end{titlepage}

@@

% TODO: Talk about how fast transfer are important for P2P backup in particular, since many replicas will need to be made and replicas will need to be moved around.
\chapter{Introduction}

@@

% TODO: Ask Sarkozy/Mihaly if they think keeping this section in is worth it.
\subsection{File Chunking} \label{sec:filechunking}
The format in which files are stored can have an effect on the efficiency of the P2P backup system. At first, backing up each file as a whole might seem like the obvious choice, since files are only useful in their entirety. However, the creators of the P2P backup systems pStore \cite{pStore} and PeerStore \cite{PeerStore} have shown that there are benefits to splitting files up into fixed-sized chunks, rather than distributing complete files to peers. In these systems, this is accomplished by creating two types of files: a "file block", which contains a section (chunk) of the original file, and a "file block list", which contains information on which file blocks compose the original file and how to order them. This allows for peer heterogeneity in terms of storage capacity. For example, if a peer needs to store a 10GB file, it does not need to find an individual peer capable of storing 10GB (which might be rare); rather, it needs to find a number of peers, each capable of storing the size of a chunk (which is magnitudes smaller in size than the original file).

@@

\subsection{Cryptography} \label{sec:crypto}
Cryptography is the study of techniques for secure communication in the presence of third parties \cite{cryptoDef}. It is heavily relied upon in the computer world to protect secrets, verify identities, and ensure data has not been tampered with. Our system uses cryptography to \textit{encrypt} users' data before replicating it to strangers' computers. Encryption is the process of obfuscating text into a state such that it cannot be understood without \textit{decryption}, the reverse process. Secret \textit{keys} are used to parameterize encryption algorithms so that, although the specific process for encryption is public, the message it encrypts remains a secret. In \textit{symmetric-key} encryption, the key used to decrypt is identical to the key used to encrypt. This differs from \textit{public-key} encryption, in which two different, yet mathematically related, keys are used.

Encryption in P2P backup systems is necessary to prevent the peers who are storing backups of data from reading it. A system in which backups were unencrypted would be much less popular with users, because they would lose privacy by using it. The terminology associated with this is \textit{confidentiality}, meaning that data that is supposed to be confidential, remains confidential. Our P2P backup system leverages BitTorrent Sync, which uses the AES encryption algorithm to encrypt backups before replicating them.
 % TODO not sure if we should describe AES here... AES is a \textit{symmetric-key} encryption algorithm, which means the same key is used to both encrypt and decrypt.

\textit{Hashing} is the process of applying a function to some arbitrary-length message, resulting in a fixed sized output, or \textit{hash}. \textit{Cryptographic hash functions} have mathematical-based properties, e.g., \textit{pre-image resistance}, that provide certain guarantees (with high confidence) about the resulting hash. Pre-image resistance refers to hash functions that cannot be \textit{reversed}. That is, given $hash(m) = h$, the hash function is pre-image resistant if an adversary cannot find the input $m$ when given the output $h$. The other required properties of cryptographic hash functions are \textit{second pre-image resistance} and \textit{collision resistance}. A hash function has second pre-image resistance if, given an input $m_1$, it is difficult to find a second input $m_2$ such that $hash(m_2) = hash(m_1)$. This is similar to collision resistance, which mandates it is difficult to find any two inputs that map to the same output.
% TODO find math to put here
% TODO why are we explaining these properties of crypto hash functions? do we even use crypto hash functions?

@@

\begin{figure} \label{fig:convergent}
  \centering
  \includegraphics[scale=0.75]{figures/ConvergentEncryption}
  \caption{Convergent Encryption from PeerStore}
\end{figure}

@@

A \textit{hash table} is a data structure that maps unique keys to values. It has attractive performance features, such as $O(1)$ average lookup time. This means that, on average, the time it takes to retrieve a value from the hash table is independent of the number of entries being stored. This performance comes from the use of a hash function when organizing the data internally. A hash table can be represented as an array of \textit{buckets}, which are slots used to store the data. Hash functions are used to map a given key to an index in the array of buckets. Thus, given a key, the hash table can compute the bucket in which the (key, value) pair is stored. In the event of hash collisions, the (key, value) pairs in each bucket can be stored as a linked list. Hash tables use hash functions that have uniform hash distribution to minimize collisions, since traversing the linked list is the bottleneck of the operation. Hash tables have $O(1)$ insertion and deletion time for the same reason.

@@

% TODO citations?

@@

\section{The BitTorrent Protocol} \label{sec:TheBitTorrentProtocol}
BitTorrent is a peer to peer protocol designed for fast file sharing. A peer who wishes to share a file creates a .torrent file containing metadata about the file to be shared. This peer is known as the file's initial \textit{seeder}, or uploader, and through the use of a BitTorrent client, starts \textit{seeding}, or uploading, the file. When a peer acts as a seeder, its sole purpose is to share the file that it is seeding with other peers. Peers who wish to download the file must acquire its .torrent file through an out-of-band channel, which is often a torrent indexer. Peers use the .torrent file to locate the initial seeder through either a centralized tracker or a DHT. These peers then connect to the initial seeder and start downloading the file. These peers are known as \textit{leechers} until they possess a full copy of the file, which is when they, themselves, become seeders. A \textit{swarm} refers to all entities, seeders and leechers, participating in a given torrent.

@@

BitTorrent can transfer files quickly due to multiple seeders each uploading different chunks of the file to each leecher. This parallelized download is enhanced by the fact that leechers, too, upload the chunks of the file they possess to other leechers who do not have those chunks. \footnote {This is a simplified version of the BitTorrent protocol. For more information, see \cite{bittorrentProtocol}}

@@

\chapter{BTBackup} \label{chap:BTBackup}

@@

% Ubuntu 12.04 x64 in VM in kvm
% Using SZTAKICloud at MTA SZTAKI in Budapest
% (how much are we allowed to talk about it? ask Mihaly)

% TODO: Is it actually called SZTAKI Cloud? I've looked through my emails, and Mihaly never calls it exactly that.
In order to test our system, we needed an environment that would allow us to create and populate our network with as many nodes as we needed for each test scenario. We used SZTAKI Cloud, a virtualization platform that runs on the open source cloud computing system \textit{OpenNebula}. SZTAKI Cloud allowed us to create and clone virtual machines for each test scenario. This is done via its web interface. % TODO: Elaborate more here, depending on how much Mihaly wants us to talk about it.

@@

Section \ref{subsec:DataRecoverySpeedfromNodesPerspective} verified our hypothesis that our system would recover data faster than an equivalent one using a direct download data transfer protocol. Our data suggests that the speed increase could be upwards of 300\%. This is possible, because the use of multiple nodes in a file recovery allows the download bandwidth of the peer to be fully utilized.

Section \ref{subsec:UploadSpeedofDataBackup} showed that our system lessens the Peer's load of backing up a file by distributing the task to the nodes as well. This has the side effect of speeding up the backup process, because nodes upload the data they have at a given time to each other, better utilizing their download bandwidth. Our experiment showed that the speed increase could be upwards of 300\%. It is important to consider the number of replicas per file here, because the replica count that is maintained affects the speed and bandwidth usage of the backup and restore procedures. As the number of replicas per file increases, more traffic must be generated to create and maintain those replicas, but the download speed when recovering will be greatly improved. Of course, the maximum recovery speed is limited by the peer's download bandwidth.

@@

\bibitem{dropboxusers} https://www.dropbox.com/news/company-info
\bibitem{onedriveusers} http://blog.onedrive.com/over-250m-people-using-skydrive/
\bibitem{ChurnResilient} S. Legtchenko, S. Monnet, P. Sens, G. Muller, Churn-Resilient Replication Strategy for Peer-to-Peer Distributed Hash-Tables
\bibitem{Kademlia} P. Maymounkov, D. Mazi\`eres, Kademlia: A Peer-to-Peer Information System Based on the XOR Metric
\bibitem{pStore} C. Batten, K. Barr, A. Saraf, S. Trepetin, pStore: A Secure Peer-to-Peer Backup System
\bibitem{PeerStore} M. Landers, H. Zhang, K. Tan, PeerStore: Better Performance by Relaxing in Peer-to-Peer Backup
\bibitem{StorageSearchP2PNetworks} J. Augustine, A. Molla, E. Morsy, G. Pandurangan, Storage and Search in Dynamic Peer-to-Peer Networks
\bibitem{p2pSurvey} A. Passarella, A survey on content-centric technologies for the current Internet: CDN and P2P solutions
\bibitem{BTSyncFAQ} http://www.bittorrent.com/help/faq/sync
\bibitem{btsynctech} http://www.bittorrent.com/sync/technology
\bibitem{btsyncuserguide} http://btsync.s3-website-us-east-1.amazonaws.com/BitTorrentSyncUserGuide.pdf
\bibitem{btsyncapi} http://www.bittorrent.com/sync/developers/api
\bibitem{cryptoDef} Rivest, Ronald L. (1990). "Cryptology". In J. Van Leeuwen. Handbook of Theoretical Computer Science 1. Elsevier.
\bibitem{bittorrentDHT} http://www.bittorrent.org/beps/bep\_0005.html
\bibitem{bittorrentProtocol} http://www.bittorrent.org/beps/bep\_0003.html
\bibitem{syslogRFC1} C. Lonvick, "The BSD syslog Protocol". https://tools.ietf.org/html/rfc3164, August 2001.
\bibitem{syslogRFC2} R. Gerhards, "The syslog Protocol". https://tools.ietf.org/html/rfc5424, March 2009.
\bibitem{skypeSupernode} S. Guha, N. Daswani, R. Jain, "An Experimental Study of the Skype Peer-to-Peer VoIP System". http://saikat.guha.cc/pub/iptps06-skype/ .
\bibitem{kademliaDHT} P. Maymounkov, D. Mazi\`eres, Kademlia: A Peer-to-peer Information System Based on the XOR Metric, IPTPS '01 Revised Papers from the First International Workshop on Peer-to-Peer Systems (2002), 53-65.
\bibitem{internetUsersByCountry} International Telecommunications Union (Geneva), Percentage of Individuals using the Internet 2000-2012, June 2013, retrieved 29 April 2014.
\bibitem{downloadSpeedsByCountry} Ookla, Household Download Index, April 2014, accessed from http://www.netindex.com/download/allcountries/ on 29 April 2014.

>>>>

% TODO make title page look nicer
\begin{titlepage}
\begin{center}
\thispagestyle{empty}
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[1.75cm]
\LARGE{\textsc {\textbf{Improving the Speed of Peer to Peer Backup Systems with BitTorrent}}}\\[0.5cm]
\vspace{0.25cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.75cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{1cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\vspace{.25cm}
\large{\textbf{Mih\'aly H\'eder}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\end{center}
\end{titlepage}

@@

\chapter{Introduction}

@@

\subsection{File Chunking} \label{sec:filechunking}
The format in which files are stored can have an effect on the efficiency of the P2P backup system. At first, backing up each file as a whole might seem like the obvious choice, since files are only useful in their entirety. However, the creators of the P2P backup systems pStore \cite{pStore} and PeerStore \cite{PeerStore} have shown that there are benefits to splitting files up into fixed-sized chunks, rather than distributing complete files to peers. In these systems, this is accomplished by creating two types of files: a "file block", which contains a section (chunk) of the original file, and a "file block list", which contains information on which file blocks compose the original file and how to order them. This allows for peer heterogeneity in terms of storage capacity. For example, if a peer needs to store a 10GB file, it does not need to find an individual peer capable of storing 10GB (which might be rare); rather, it needs to find a number of peers, each capable of storing the size of a chunk (which is magnitudes smaller in size than the original file).

@@

\subsection{Cryptography} \label{sec:crypto}
Cryptography is the study of techniques for secure communication in the presence of third parties \cite{cryptoDef}. It is heavily relied upon in the computer world to protect secrets, verify identities, and ensure data has not been tampered with. Our system uses cryptography to \textit{encrypt} users' data before replicating it to strangers' computers. Encryption is the process of obfuscating text into a state such that it cannot be understood without \textit{decryption}, the reverse process. Secret \textit{keys} are used to parameterize encryption algorithms so that, although the specific process for encryption is public, the message it encrypts remains a secret. In \textit{symmetric-key} encryption, the key used to decrypt is identical to the key used to encrypt. This differs from \textit{public-key} encryption, in which two different, yet mathematically related, keys are used \cite{ModernCrypto}.

Encryption in P2P backup systems is necessary to prevent the peers who are storing backups of data from reading it. A system in which backups were unencrypted would be much less popular with users, because they would lose privacy by using it. The term associated with this is \textit{confidentiality}, meaning that data that is supposed to be confidential, remains confidential \cite{OWASPCrypto}. Our P2P backup system leverages BitTorrent Sync, which uses the AES encryption algorithm to encrypt backups before replicating them.
 % TODO not sure if we should describe AES here... AES is a \textit{symmetric-key} encryption algorithm, which means the same key is used to both encrypt and decrypt.

\textit{Hashing} is the process of applying a function to some arbitrary-length message, resulting in a fixed sized output, or \textit{hash}. \textit{Cryptographic hash functions} have mathematical-based properties, e.g., \textit{pre-image resistance}, that provide certain guarantees (with high confidence) about the resulting hash. Pre-image resistance refers to hash functions that cannot be \textit{reversed}. That is, given $hash(m) = h$, the hash function is pre-image resistant if an adversary cannot find the input $m$ when given the output $h$. The other required properties of cryptographic hash functions are \textit{second pre-image resistance} and \textit{collision resistance}. A hash function has second pre-image resistance if, given an input $m_1$, it is difficult to find a second input $m_2$ such that $hash(m_2) = hash(m_1)$. This is similar to collision resistance, which describes hash functions for which it is difficult to find any two inputs that map to the same output \cite{CryptoHashFunctions}.
% TODO find math to put here
% TODO why are we explaining these properties of crypto hash functions? do we even use crypto hash functions?

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.75]{figures/ConvergentEncryption}
  \caption{Convergent Encryption from PeerStore \label{fig:convergent}}
\end{figure}

@@

A \textit{hash table} is a data structure that maps unique keys to values. It has attractive performance features, such as $O(1)$ average lookup time. This means that, on average, the time it takes to retrieve a value from the hash table is independent of the number of entries being stored. This performance comes from the use of a hash function when organizing the data internally. A hash table can be represented as an array of \textit{buckets}, which are slots used to store the data. Hash functions are used to map a given key to an index in the array of buckets. Thus, given a key, the hash table can compute the bucket in which the (key, value) pair is stored. In the event of hash collisions, the (key, value) pairs in each bucket can be stored as a linked list. Hash tables use hash functions that have uniform hash distribution to minimize collisions, since traversing the linked list is the bottleneck of the operation. Hash tables have $O(1)$ insertion and deletion time for the same reason \cite{HashTable}.

@@

@@

\section{The BitTorrent Protocol} \label{sec:TheBitTorrentProtocol}
BitTorrent is a peer to peer protocol designed for fast file sharing \cite{bittorrentProtocol}. A peer who wishes to share a file creates a .torrent file containing metadata about the file to be shared. This peer is known as the file's initial \textit{seeder}, or uploader, and through the use of a BitTorrent client, starts \textit{seeding}, or uploading, the file. When a peer acts as a seeder, its sole purpose is to share the file that it is seeding with other peers. Peers who wish to download the file must acquire its .torrent file through an out-of-band channel, which is often a torrent indexer. Peers use the .torrent file to locate the initial seeder through either a centralized tracker or a DHT. These peers then connect to the initial seeder and start downloading the file. These peers are known as \textit{leechers} until they possess a full copy of the file, which is when they, themselves, become seeders. A \textit{swarm} refers to all entities, seeders and leechers, participating in a given torrent.

@@

BitTorrent can transfer files quickly due to multiple seeders each uploading different chunks of the file to each leecher. This parallelized download is enhanced by the fact that leechers, too, upload the chunks of the file they possess to other leechers who do not have those chunks.\footnote{This is a simplified version of the BitTorrent protocol. For more information, see \cite{bittorrentProtocol}}

@@

\chapter{BTBackup Design} \label{chap:BTBackup}

@@

In order to test our system, we needed an environment that would allow us to create and populate our network with as many nodes as we needed for each test scenario. We used SZTAKI Cloud, a virtualization platform that runs on the open source cloud computing system \textit{OpenNebula}. SZTAKI Cloud allowed us to create and clone virtual machines for each test scenario.

@@

Section \ref{subsec:DataRecoverySpeedfromNodesPerspective} verified our hypothesis that our system would recover data faster than an equivalent one using a direct download data transfer protocol. Our data suggests that the speed increase could be upwards of 300\%. This is possible, because the use of multiple nodes in a file recovery allows the download bandwidth of the peer to be fully utilized. This has the side effect of distributing the load of backup recovery among the nodes.

Section \ref{subsec:UploadSpeedofDataBackup} showed that our system lessens the Peer's load of backing up a file by distributing the task to the nodes as well. This has the side effect of speeding up the backup process, because nodes upload the data they have at a given time to each other, better utilizing their download bandwidth. Due to the recurring need for replica replacement, the benefits of speeding up the backup process will be experienced throughout the backup's lifetime. Our experiment showed that the speed increase could be upwards of 300\%. It is important to consider the number of replicas per file here, because the replica count that is maintained affects the speed and bandwidth usage of the backup and restore procedures. As the number of replicas per file increases, more traffic must be generated to create and maintain those replicas, but the download speed when recovering will be greatly improved. Of course, the maximum recovery speed is limited by the peer's download bandwidth.

@@

\bibitem{dropboxusers} https://www.dropbox.com/news/company-info
\bibitem{onedriveusers} http://blog.onedrive.com/over-250m-people-using-skydrive/
\bibitem{ChurnResilient} S. Legtchenko, S. Monnet, P. Sens, G. Muller, Churn-Resilient Replication Strategy for Peer-to-Peer Distributed Hash-Tables
\bibitem{Kademlia} P. Maymounkov, D. Mazi\`eres, Kademlia: A Peer-to-Peer Information System Based on the XOR Metric
\bibitem{pStore} C. Batten, K. Barr, A. Saraf, S. Trepetin, pStore: A Secure Peer-to-Peer Backup System
\bibitem{PeerStore} M. Landers, H. Zhang, K. Tan, PeerStore: Better Performance by Relaxing in Peer-to-Peer Backup
\bibitem{StorageSearchP2PNetworks} J. Augustine, A. Molla, E. Morsy, G. Pandurangan, Storage and Search in Dynamic Peer-to-Peer Networks
\bibitem{p2pSurvey} A. Passarella, A survey on content-centric technologies for the current Internet: CDN and P2P solutions
\bibitem{BTSyncFAQ} http://www.bittorrent.com/help/faq/sync
\bibitem{btsynctech} http://www.bittorrent.com/sync/technology
\bibitem{btsyncuserguide} http://btsync.s3-website-us-east-1.amazonaws.com/BitTorrentSyncUserGuide.pdf
\bibitem{btsyncapi} http://www.bittorrent.com/sync/developers/api
\bibitem{cryptoDef} Rivest, Ronald L. (1990). "Cryptology". In J. Van Leeuwen. Handbook of Theoretical Computer Science 1. Elsevier.
\bibitem{CryptoHashFunctions} P. Rogaway, T. Shrimpton. "Cryptographic Hash-Function Basics: Definitions, Implications, and Separations for Preimage Resistance, Second-Preimage Resistance, and Collision Resistance". 2004. 
\bibitem{ModernCrypto} N. P. Smart. "Cryptography, An Introduction : Third Edition". 10 April 2013. Retrieved from http://www.cs.bris.ac.uk/~nigel/Crypto\_Book/ .
\bibitem{OWASPCrypto} OWASP. "Guide to Cryptography". 25 February 2012. Retrieved on 30 April 2014 from https://www.owasp.org/index.php/Guide\_to\_Cryptography .
\bibitem{HashTable} Donald Knuth. "The Art of Computer Programming, Volume 3: Sorting and Searching (2nd ed.)". 1998. Addison-Wesley, pp. 513–558. ISBN 0-201-89685-0.
\bibitem{bittorrentDHT} http://www.bittorrent.org/beps/bep\_0005.html
\bibitem{bittorrentProtocol} http://www.bittorrent.org/beps/bep\_0003.html
\bibitem{syslogRFC1} C. Lonvick, "The BSD syslog Protocol". https://tools.ietf.org/html/rfc3164, August 2001.
\bibitem{syslogRFC2} R. Gerhards, "The syslog Protocol". https://tools.ietf.org/html/rfc5424, March 2009.
\bibitem{skypeSupernode} S. Guha, N. Daswani, R. Jain, "An Experimental Study of the Skype Peer-to-Peer VoIP System". http://saikat.guha.cc/pub/iptps06-skype/ .
\bibitem{kademliaDHT} P. Maymounkov, D. Mazi\`eres, Kademlia: A Peer-to-peer Information System Based on the XOR Metric, IPTPS '01 Revised Papers from the First International Workshop on Peer-to-Peer Systems (2002), 53-65.
\bibitem{internetUsersByCountry} International Telecommunications Union (Geneva), Percentage of Individuals using the Internet 2000-2012, June 2013, retrieved 29 April 2014.
\bibitem{downloadSpeedsByCountry} Ookla, Household Download Index, April 2014, accessed from http://www.netindex.com/download/allcountries/ on 29 April 2014.
